# Functional programming in R

This tutorial will cover the basics of functional programming in R using the purrr package. In functional-style programming, we can break big problems into smaller pieces, and then solve each piece with a function. This is a powerful way to write code because it is more modular, easier to read, and less error-prone. This material is based off of Hadley Wicham's Advanced R book, which is a great resource for learning more about functional programming in R (https://adv-r.hadley.nz/).

Let's start by installing and loading the libraries we will need for this tutorial:
```{r}
install_and_load <- function(package) {
  if (!require(package, character.only = TRUE)) {
    install.packages(package, dependencies = TRUE)
  }
  library(package, character.only = TRUE)
  return()
}

packages <- c("purrr", "CoordinateCleaner", "dplyr", "tidyr", "rgbif")
lapply(packages, install_and_load) # Will return NULL if successful
```

## 1. Functionals

The first concept we will cover is functionals. Functionals are functions that take a function as an input and return a vector (or list) as output.

Examples: `lapply()`, `apply()`, `tapply()`, or `purrr::map()`

Functionals are commonly used as an alternative to for loops. For loops aren't bad, but they are often  “too flexible” (unclear what you are producing, potential side effects, not input/output oriented, etc.)

We will be using the purrr package to work with functionals. Purrr provides a consistent and user friendly set of tools for working with functions. The most basic purrr function is `map()`. This function takes a vector or list as input and applies a function to each element. The output is a list where each element is the result of applying the function to the corresponding element of the input.

```{r}
# here is a simple example of using map to print each element of a vector
map(c("hi", "hello", "yoohoo"), function(greeting) print(greeting))

# you can also use the shorthand ~ to define a function
map(c("hi", "hello", "yoohoo"), ~print(.x))

# or you can use an existing function, if there is only one argument
map(c("hi", "hello", "yoohoo"), print)
```

How is this different from a non-functional? Let's say you wanted to simulate rolling die with different number of sides. You could use a for loop to iterate over each number of sides:
```{r}
d6_roll <- numeric(10)
d10_roll <- numeric(10)
magic8ball_roll <- numeric(10)
for (i in 1:10) {
  d6_roll[i] <- sample(1:6, 1, replace = TRUE)
  d10_roll[i] <- sample(1:10, 1, replace = TRUE)
  magic8ball_roll[i] <- sample(1:20, 1, replace = TRUE)
}
d6_roll
d10_roll
magic8ball_roll
```

Or you could define a function to roll a die given some number of sides and then use map to run that function 10 times:
```{r}
roll <- function(sides) sample(1:sides, 1, replace = TRUE)

d6_roll <- map_dbl(1:10, ~roll(sides = 6))
d10_roll <- map_dbl(1:10, ~roll(sides = 10))
magic8ball_roll <- map_dbl(1:10, ~roll(sides = 20))

d6_roll
d10_roll
magic8ball_roll
```
What is nice about the second approach is that it is more modular and easier to read. You can see that we are rolling a die with 6 sides, 10 sides, and 20 sides, and we are doing it 10 times. ALso, if there happened to be a bug in the roll function, we could fix it in one place and it would be fixed for all the rolls. We also won't have any side effects from the for loop, such as accidentally overwriting a variable.

Note that `map_dbl()` is used here instead of `map()` because we know that the output of roll is a number (known as a double) and we want to simplify the output to a vector rather than a list (as would be returned by map). You can also use `map_chr()` if you know the output will be a character vector, `map_lgl()` if you know the output will be a logical vector, etc.

If you don't like the tidyverse, you can also use `sapply()` or `lapply()` to do the same thing:
```{r}
sapply(1:10, function(x) roll(sides = 6))
```

**Example 1: Using `map()` to pull occurrence data for a list of species**

Now let's try applying `map()` to a more realistic research problem. Imagine we have a list of species and we want to pull occurrence data for each species from the GBIF database. We can use a functional approach with the `map()` function to iterate over each species and pull the data.

```{r}
# First, let's define a list of species we want to pull occurrence data for
# Let's say we are working on three different projects and each project is focused on a different species
species <- 
  c(
    Project01 = "Sceloporus occidentalis", 
    Project02 = "Phrynosoma blainvillii", 
    Project03 = "Plestiodon skiltonianus"
  )

# Map over the species vector and use occ_search to pull occurrence data for each species
#  ~ .x is a shorthand for function(x) x
# .progress = TRUE will show a progress bar (super handy for the impatient, like me)
occ_records <- map(species, ~occ_search(scientificName = .x, hasCoordinate = TRUE, limit = 1000), .progress = TRUE)

# This will provide a list where each element is the occurrence data for a species
# For example, let's look at the Project01 element
# Note that the names of the list correspond to the names of the species vector 
occ_records[["Project01"]]

# This element is an rgbif object
class(occ_records[["Project01"]])
# And therefore has several different elements itself
names(occ_records[["Project01"]])

# What we care about is the "data" element, and we can use map to pull just that element out
occ_data <- map(occ_records, "data")

# We can also bind the rows of our list together to make a dataframe
# The .id argument will add a column with the name of the list element, in this case our original project ID (this is super handy for keeping track of where things come from in a list)
occ_df <- bind_rows(occ_data, .id = "project")
head(occ_df)
```

**Example 2: the wonders of `pmap()`**

Now let's say you want to download occurrence records for specimens from the MVZ, LACM, and CAS museums. You could do a nested for-loop to iterate over each species and each museum...or you could use `pmap()` to iterate over all your combinations at once.

```{r}
museums <- c("MVZ", "LACM", "CAS")

# To make our code a little easier to read and put the "function" in functional, let's first make a function that will pull occurrence data for a species from a museum
occ_search_museum <- function(species, museum) {
  occ_search(
    scientificName = species, 
    hasCoordinate = TRUE, 
    limit = 1000,
    basisOfRecord = "PRESERVED_SPECIMEN",
    institutionCode = museum
  )$data
} 

# Now, lets create grid of all the combinations of species and museums
# This will be a dataframe with two columns, one for species and one for museum with all possible combinations of the two 
species_museum <- expand_grid(speciesID = species, museumID = museums)
print(species_museum)

# Instead of map() we will now use pmap(), which will take each row of species_museum to fill in the arguments of occ_search_museum()
# Notice that we use \(x, y) to define the arguments of the function, this is the same thing as ~function(x, y), but in a more concise format
museum_records <- 
  pmap(species_museum, \(speciesID, museumID) occ_search_museum(species = speciesID, museum = museumID), .progress = TRUE) %>%
  bind_rows()
# We could have also used pmap(species_museum, occ_search_museum) since our arguments are in order, but this way is a little more explicit and less prone to bugs
```

## 2. Functional factories

Another key concept in functional programming is functional factories. Functional factories are functions that return other functions. This can be useful if you have a lot of different functions that all need to use the same arguments.

**Example 3: creating a museum factory**

Below we use a factory function to create functions that will pull occurrence data from a specific museum.

```{r}
museum_factory <- function(museum){
  occ_search_museum <- function(species) {
    result <- occ_search(
      scientificName = species, 
      hasCoordinate = TRUE, 
      limit = 1000,
      basisOfRecord = "PRESERVED_SPECIMEN",
      institutionCode = museum
    )$data
  }
}

# We can use our factory to create functions that will pull occurrence data for a species from a specific museum
search_mvz <- museum_factory("MVZ")
search_lacm <- museum_factory("LACM")
search_cas <- museum_factory("CAS")

# We can use our new functions to pull occurrence data from a specific museum
search_mvz("Crotalus cerastes")
```

## 3. Function operators

Our final functional programming technique is the use of function operators. Function operators are functions that take other functions as input and return a new function as output. In other words, they are just function factories that take a function as input. Function operators can be particularly useful for error handling, as you will see in the next example.

**Example 4: catching errors with safely() and possibly()**

Have you ever been runnning a super long loop and then it fails 90% of the way through? Argh! Wouldn't it be better if it stored the results of everything that worked and then told you what went wrong? Well, that is what `safely()` does! `safely()` will take in your function and return a new function that will return a list with the result of the function if it works or an error message if the function fails.

```{r}
# Let's imagine we have an invalid species name in our list
species <- 
  c(
    Project01 = "Phrynosoma blainvillii", 
    Project02 = "Plestiodon skiltonianus",
    Project03 = "Bogus bogus"
  )

# We won't initially get an error from our search function because the function will just return a NULL
mvz_specimens <- map(species, search_mvz, .progress = TRUE)

# But if we try do stuff with the list, such as pull coordinates out, we will get an error
pull_coords <- function(data){
  data <- dplyr::select(data, species, decimalLatitude, decimalLongitude)
  return(data)
}
map(mvz_specimens, pull_coords) # this will fail!

# We can use safely() to create a new function that will catch any errors
safe_function <- safely(pull_coords)
coords <- map(mvz_specimens, safe_function)

# If we look at the output, we will see that the output is a list with two elements, result and error
coords[["Project01"]]

# If the function was successful , the result element will contain the output of the function and the error element will be NULL
coords[["Project01"]]$result
coords[["Project01"]]$error

# If the function failed, the result element will be NULL and the error element will contain the error message
coords[["Project03"]]$result
coords[["Project03"]]$error

# You can use map() to pull out the error messages
map(coords, "error")

# You can also use map to pull out the elements with no errors and compact() to remove any NULL elements
coords <- map(coords, "result") %>% compact()
```

Another handy function operator is `possibly()`. This will return a list with the result or NULL if the function failed. This can be useful if you know that some of your functions will fail, but you just want to ignore them and keep going and then drop them later.
```{r}
# Possible_function will return NULL if the function fails
# You can change the otherwise argument to return something else (e.g., NA or "No result"), if you want
possible_function <- possibly(pull_coords, otherwise = NULL)

# Now we can use possible_function() to pull out the coordinates and ignore any errors
coords <- map(mvz_specimens, possible_function)

# And then we can use compact() to remove any NULL elements
coords <- compact(coords)
```

# Try it out!

Now it's time for you to be functional! The following code uses the CoordinateCleaner package to clean up some occurrence data using the `clean_coordinates()`function; this function takes in a dataframe of occurrence data and a list of tests to run. The output is a dataframe with a column for each test that was run and a flag for each record that failed that test. Rewrite the  code in a functional programming style using the purrr functions we have learned about:
```{r}
library(CoordinateCleaner)

clean_project1 <- clean_coordinates(occ_data[["Project01"]], tests = c("institutions", "seas", "duplicates", "outliers", "urban"))
clean_project2 <- clean_coordinates(occ_data[["Project02"]], tests = c("institutions", "outliers", "seas", "duplicates", "urban"))
clean_project3 <- clean_coordinates(occ_data[["Project03"]], tests = c("institutions", "outliers", "seas", "duplicates", "urban"))

clean_list <- list(Project01 = clean_project1, Project02 = clean_project2, Project03 = clean_project3)
```

The tests argument specifies which tests to run. The tests specified here are:
- **institutions** tests a radius around known biodiversity institutions from institutions. The radius is inst_rad.
- **outliers** tests each species for outlier records. Depending on the outliers_mtp and outliers.td arguments either flags records that are a minimum distance away from all other records of this species (outliers_td) or records that are outside a multiple of the interquartile range of minimum distances to the next neighbour of this species (outliers_mtp). Three different methods are available for the outlier test: "If “outlier” a boxplot method is used and records are flagged as outliers if their mean distance to all other records of the same species is larger than mltpl * the interquartile range of the mean distance of all records of this species. If “mad” the median absolute deviation is used. In this case a record is flagged as outlier, if the mean distance to all other records of the same species is larger than the median of the mean distance of all points plus/minus the mad of the mean distances of all records of the species * mltpl. If “distance” records are flagged as outliers, if the minimum distance to the next record of the species is > tdi.
- **seas** tests if coordinates fall into the ocean.
- **validity** checks if coordinates correspond to a lat/lon coordinate reference system. This test is always on, since all records need to pass for any other test to run.
**For a list of all available tests see the function documentation with ?clean_coordinates.**

```{r}
# Write your code here
```

Below is code to plot the results of the tests for project 1. The .summary column contains the overall result of the tests. The other columns contain the results of the individual tests. If the test failed, the value in the column will be FALSE and if the test passed, the value will be TRUE. If any test failed, the value in the .summary column will be FALSE.

```{r}
problems <- 
  clean_project1 %>%
  pivot_longer(starts_with("."), names_to = "test", values_to = "result") %>%
  mutate(test = factor(test, levels = c(setdiff(unique(test), ".summary"), ".summary")))

ggplot(problems) +
  geom_point(aes(x = decimalLongitude, y = decimalLatitude, color = result)) +
  facet_wrap(~test, nrow = 1) +
  coord_quickmap()
```

Write a function that will plot the results of the tests for a given project. The function should take a cleaned dataframe as input and returned a ggplot object. 

```{r}
# Write your code here
```

After reviewing your results, you want to exclude any points that are in the ocean, but ignore all of the other flags. Use purrr to filter out any points that are flagged as "seas" in the `clean_list` object.

```{r}
# Write your code here
```